Global code motion of congruent instructions

### 2017-11-05 cgo/paper.tex
correction from reviewers

We have implemented the pass in \LLVM{}. It is based on Global Value Numbering
infrastructure available in \LLVM{}. The experimental results show an average
saving of 1\% on the total compilation time on \SPEC{}. \GCM{} enables more
inlining and exposes more loop invariant code motion opportunities in majority
of the benchmarks. We have also seen execution time improvements in a few of
\SPEC{} benchmarks viz. mcf and sjeng, moreover, register spills reduced by 2\%
on the \SPEC{} benchmarks suite when compiled for \xlinux{}. \GCM{} is an
optimistic algorithm in the sense that it considers all congruent instructions
in a function as potential candidates. We make an extra effort to hoist
candidates by partitioning the potential candidates in a way to enable partial
hoisting in case a common hoisting point for all the candidates cannot be
found. We also formalize why register pressure reduces as a result of \gcm{},
and how \gcm{} increases instruction level parallelism thereby enabling more
\ooo{} execution on modern architectures.
\end{abstract}

@@

There has been a lot of work both in industry and academia to hoist and sink
code out of branches, and in general global scheduling
\cite{click1995global}. Some relate code-hoisting to code-size optimization
\cite{rosen1988global} and many \cite{barany2013, shobaki2013} use global
scheduling to improve performance. Most of the recent work on global scheduling
are done using integer linear programming (ILP,) which results in prohibitively
high compile time. To the best of our knowledge we have not found any reference
which explored \gcm{} of congruent instructions in as much detail us. A part of
our implementation (aggressive code hoisting) is already merged
in \LLVM{} trunk, however, a general implementation of \gcm{} of congruent (non
redundant) expressions is still missing from \GCC{} and \LLVM{} trunk. The main
contributions of this paper are:
\begin{itemize}[leftmargin=*,topsep=0pt]
\item a new optimistic algorithm to move congruent instructions out of branches;
\item a cost model to reduce interference and hence, reduce spills;
\item a technique to maximize hoisting in an optimistic approach by partitioning
  the list of potential candidates sorted by their DFS visit number;
\item experimental evaluation of our implementation in \LLVM{} which combines
  \SSA{} based liveness analysis, and ranking of expressions to move very busy
  expressions in order to reduce register spills and improve performance in some
  cases.
\end{itemize}

@@

We compute merge sets, of the control flow graph, which remain same throughout
the \GCM{} transformation. For \GCM{} we only want to know if a use operand is a
kill (to compute changes in register pressure). For that we only need to know
whether the use is also required later in the execution path. A variable is live
out of a basic-block B if it is used in the merge set of B. We compute the
live-out relation on-demand when profitability of hoist/sink candidates is to be
evaluated. A simplified version of \emph{isLiveOutUsingMergeSet} is presented
here:

@@

\subsubsection{Legality of hoisting scalars}
Scalars are the easiest to hoist because we do not have to analyze them for
aliasing memory references. As long as all the operands are available, or can be
made available by rematerialization,) the scalar computations can be hoisted.

@@

Hoisting loads/stores across calls also require precise analysis of all the
memory addresses accessed by the call. The current implementation being an
intraprocedural pass, cannot hoist aggressively across calls. In the presence of
pure calls, loads can be hoisted but stores can't. Also, if a call throws
exceptions, or if it may not return, nothing can be hoisted across that call.

@@

\subsubsection{Barriers}
Barriers are based on the concept of pinned instructions \cite{click1995global}
but extended to adapt to \LLVM{} IR. Since a basic-block in \LLVM{} IR is
actually an extended basic-block because there might be calls in the middle of a
basic-block which might not return. Essentially, any instruction that cannot
guarantee progress is marked as a barrier. In the absence of context, as in our
current implementation, some instructions which might be safe are still
classified as barriers, e.g., volatile loads/stores, calls with missing
attributes.

@@

\subsubsection{Reduce register pressure}
\label{hoist:reg-pressure}
Hoisting upwards will decrease the live-range of its use, if it is a last use (a
kill,) but increase the live-range of its definition. Conversely, sinking will
decrease the live-range of the defined register but increase the live-range for
killed operands. If the live-range after \GCM{} will decrease, it will be
moved. Essentially, as long as there is one killed operand, code hoisting will
either decrease or preserve the register pressure.  Similarly, code-sinking will
either decrease or preserve the register pressure as long as there is one
operand killed at most.  The following example explains how \GCM{} of can reduce
the register pressure.

@@

\subsubsection{Partitioning the list of hoist candidates to maximize hoisting}
\label{subsec:partition}
%The partition algorithm is described as follows:
In the approach, described in Section~\ref{subsec:optimistic}, it is possible
that a common hoisting point for all the instructions does not satisfy legality
or profitability checks. In these cases, it is still possible to `partially'
hoist a subset of instructions by splitting the set of candidates and finding a
closer hoisting point for each subset.

In order to hoist a subset of identical instructions, we partition the list of
all candidates in a way to maximize the total number of hoist instructions.  By
sorting the list of all the candidates in the increasing order of their depth
first search discovery time stamp \cite{clrs} (DFSIn numbers,) we make sure that
candidates closer in the list have their common dominator nearby. In
Figure-\ref{fig:dfsin}, if B3, B4, B5, and B6 have identical computations and if
for some reason, they cannot be hoisted at B0, then we partition the set of
hoistable candidates in their DFSIn order. In this case the DFSIn ordering would
be B3, B4, B5, B6 which will allow the instructions in (B3, B4) to be hoisted at
B1 and those in (B5, B6) to be hoisted at B2. Even if instructions in (B3, B4)
could not be hoisted, for some reason, to B1, the instructions in (B5, B6) can
still be hoisted - if legal and profitable - to B2.

@@

% TODO: Rewrite the following.
In our current implementation we keep as many candidates in one set as possible
(greedy approach.) To start with, first two candidates are tried for
hoistability (see Section~\ref{subsec:optimistic},) if that passes then the next
instruction is tried (if available) and so on until a non-hoistable instruction
is found. The partition starts from that point.  We split the list at a point
where the legality checks fail to hoist subset of candidates which are legal to
hoist and then start finding new hoisting point for the remaining ones.

@@

We see some nice gains in code size at -Os with \gcm{} (403.gcc, 447.dealII, and
483.xalancbmk) as shown in Table~\ref{tab:code-size}.  The code size listed here
is the text size delta (from linux size command) of the final executable for
each benchmark.  On the other hand, some benchmarks like 473.astar increased in
code size by almost 2\%. This is because once the code-size of a function
decreases (due to \GCM{},) it becomes cheaper to inline. As we can see in
Table~\ref{tab:stats}, four more functions got inlined in 473.astar. Except for
447.dealII, 450.soplex, 482.sphinx3 and 483.xalancbmk all other benchmarks got
more functions inlined. Also, since the pass runs early, it affects many
optimizations which rely on the number of instructions, length of the use-def
chain, and other metrics.

@@

\begin{table}[h!]
  \begin{center}
    \scalebox{0.7}{
    \begin{tabular}{|l|r|r|r|}
      \hline
Number of spills  &base(-Ofast) &\GCM{}(-Ofast) &\%change\\\hline
400.perlbench	  & 2539	& 2493	& 0.98 \\\hline
401.bzip2	  & 718	        & 707	& 0.98 \\\hline
403.gcc	          & 5782	& 5722	& 0.99 \\\hline
429.mcf	          & 14	        & 17	& 1.21 \\\hline
433.milc	  & 635	        & 642	& 1.01 \\\hline
444.namd	  & 3223	& 3274	& 1.01 \\\hline
445.gobmk	  & 2166	& 2173	& 1.00 \\\hline
447.dealII	  & 11074	& 10234	& 0.92 \\\hline
450.soplex	  & 1122	& 1125	& 1.00 \\\hline
453.povray	  & 4701	& 4692	& 1.09 \\\hline
456.hmmer	  & 1197	& 1274	& 1.06 \\\hline
458.sjeng	  & 168	        & 176	& 1.05 \\\hline
462.libquantum	  & 118	        & 118	& 1.00 \\\hline
464.h264ref	  & 3379	& 3454	& 1.02 \\\hline
470.lbm	          & 33	        & 33	& 1.00 \\\hline
471.omnetpp	  & 524	        & 527	& 1.00 \\\hline
473.astar	  & 180	        & 201	& 1.12 \\\hline
482.sphinx3	  & 674	        & 670	& 0.99 \\\hline
483.xalancbmk	  & 5102	& 4965	& 0.97 \\\hline
Grand Total	  & 43349	& 42497	& 0.98 \\\hline
    \end{tabular}}
  \end{center}
  \caption{Number of spills before and after \gcm{}(\GCM{}) on SPEC2006 at
    -Ofast. Lower is better in \%age change. Spills reduced by 2\%.}
  \label{tab:spills}
\end{table}

Spills are listed in Table~\ref{tab:spills}. Spills were reduced by an average
of 2\% on the \SPEC{} benchmark suite. This is a result of cost-model which
limits the \gcm{} to those candidates which do not increase the register
pressure (except for loads Section~\ref{subsec:cost-models},) overall spills
still decreases.

>>>>

We have implemented the pass in \LLVM{}. It is based on Global Value Numbering
infrastructure available in \LLVM{}. The experimental results show an average
saving of 1\% on the total compilation time on \SPEC{}. \GCM{} enables more
inlining and exposes more loop invariant code motion opportunities in the
majority of benchmarks. We have also seen execution time improvements in a few
of \SPEC{} benchmarks viz. mcf and sjeng, moreover, register spills reduced by
2\% on the \SPEC{} benchmarks suite when compiled for \xlinux{}. \GCM{} is an
optimistic algorithm in the sense that it considers all congruent instructions
in a function as potential candidates. We make an extra effort to hoist
candidates by partitioning the potential candidates in a way to enable partial
hoisting in case a common hoisting point for all the candidates cannot be
found. We also formalize why register pressure reduces as a result of \gcm{},
and how \gcm{} increases instruction level parallelism thereby enabling more
\ooo{} execution on modern architectures.
\end{abstract}

@@

There has been a lot of work both in industry and academia to hoist and sink
code out of branches, and in general global scheduling
\cite{click1995global}. Some relate code-hoisting to code-size optimization
\cite{rosen1988global} and many \cite{barany2013, shobaki2013} use global
scheduling to improve performance. Most of the recent work on global scheduling
are done using integer linear programming (ILP) which results in prohibitively
high compile time. To the best of our knowledge we have not found any reference
which explored \gcm{} of congruent instructions in as much detail us. A part of
our implementation (aggressive code hoisting) is already merged
in \LLVM{} trunk, however, a general implementation of \gcm{} of congruent (non
redundant) expressions is still missing from \GCC{} and \LLVM{} trunk. The main
contributions of this paper are:
\begin{itemize}[leftmargin=*,topsep=0pt]
\item a new optimistic algorithm to move congruent instructions out of branches;
\item a cost model to reduce interference and hence, reduce spills;
\item a technique to maximize hoisting in an optimistic approach by partitioning
  the list of potential candidates sorted by their DFS visit number;
\item experimental evaluation of our implementation in \LLVM{} which combines
  \SSA{} based liveness analysis, and ranking of expressions to move very busy
  expressions in order to reduce register spills and improve performance in some
  cases.
\end{itemize}

@@

We compute merge sets, of the control flow graph, which remains the same
throughout the \GCM{} transformation. For \GCM{} we only want to know if a use
operand is a kill (to compute changes in register pressure). For that we only
need to know whether the use is also required later in the execution path. A
variable is live out of a basic-block B if it is used in the merge set of B. We
compute the live-out relation on-demand when profitability of hoist/sink
candidates is to be evaluated. A simplified version of
\emph{isLiveOutUsingMergeSet} is presented here:

@@

\subsubsection{Legality of hoisting scalars}
Scalars are the easiest to hoist because we do not have to analyze them for
aliasing memory references. As long as all the operands are available, or can be
made available by rematerialization, the scalar computations can be hoisted.

@@

Hoisting loads/stores across calls also requires precise analysis of all the
memory addresses accessed by the call. The current implementation being an
intraprocedural pass, cannot hoist aggressively across calls. In the presence of
pure calls, loads can be hoisted but stores can't. Also, if a call throws
exceptions, or if it may not return, nothing can be hoisted across that call.

@@

\subsubsection{Barriers}
Barriers are based on the concept of pinned instructions \cite{click1995global}
but extended to adapt to \LLVM{} IR. A basic-block in \LLVM{} IR is actually an
extended basic-block because there might be non-returning calls in the middle of
a basic-block. Essentially, any instruction that cannot guarantee progress is
marked as a barrier. In the absence of context, as in our current
implementation, some instructions which might be safe are still classified as
barriers, e.g., volatile loads/stores, calls with missing attributes.

@@

\subsubsection{Reduce register pressure}
\label{hoist:reg-pressure}
Hoisting upwards will decrease the live-range of its use, if it is a last use (a
kill) but increase the live-range of its definition. Conversely, sinking will
decrease the live-range of the defined register but increase the live-range for
killed operands. If the live-range after \GCM{} will decrease, it will be
moved. Essentially, as long as there is one killed operand, code hoisting will
either decrease or preserve the register pressure.  Similarly, code-sinking will
either decrease or preserve the register pressure as long as there is one
operand killed at most.  The following example explains how \GCM{} of can reduce
the register pressure.

@@

\subsubsection{Partitioning the list of hoist candidates to maximize hoisting}
\label{subsec:partition}
%The partition algorithm is described as follows:
In the approach described in Section~\ref{subsec:optimistic}, it is possible
that a common hoisting point for all the instructions does not satisfy legality
or profitability checks. In these cases, it is still possible to `partially'
hoist a subset of instructions by splitting the set of candidates and finding a
closer hoisting point for each subset.

In order to hoist a subset of identical instructions, we partition the list of
all candidates in a way to maximize the total number of hoist instructions.  By
sorting the list of all the candidates in the increasing order of their depth
first search discovery time stamp \cite{clrs} (DFSIn numbers) we make sure that
candidates closer in the list have their common dominator nearby. In
Figure-\ref{fig:dfsin}, if B3, B4, B5, and B6 have identical computations and if
for some reason, they cannot be hoisted at B0, then we partition the set of
hoistable candidates in their DFSIn order. In this case the DFSIn ordering would
be B3, B4, B5, B6 which will allow the instructions in (B3, B4) to be hoisted at
B1 and those in (B5, B6) to be hoisted at B2. Even if instructions in (B3, B4)
could not be hoisted, for some reason, to B1, the instructions in (B5, B6) can
still be hoisted - if legal and profitable - to B2.

@@

% TODO: Rewrite the following.
In our current implementation we keep as many candidates in one set as possible
(greedy approach.) To start with, the first two candidates are tried for
hoistability (see Section~\ref{subsec:optimistic}) if that passes then the next
instruction is tried (if available) and so on until a non-hoistable instruction
is found. The partition starts from that point.  We split the list at a point
where the legality checks fail to hoist subset of candidates which are legal to
hoist and then start finding new hoisting point for the remaining ones.

@@

We see some nice gains in code size at -Os with \gcm{} (403.gcc, 447.dealII, and
483.xalancbmk) as shown in Table~\ref{tab:code-size}.  The code size listed here
is the text size delta (from linux size command) of the final executable for
each benchmark.  On the other hand, some benchmarks like 473.astar increased in
code size by almost 2\%. This is because once the code-size of a function
decreases (due to \GCM{}) it becomes cheaper to inline. As we can see in
Table~\ref{tab:stats}, four more functions got inlined in 473.astar. Except for
447.dealII, 450.soplex, 482.sphinx3 and 483.xalancbmk all other benchmarks got
more functions inlined. Also, since the pass runs early, it affects many
optimizations which rely on the number of instructions, length of the use-def
chain, and other metrics.

@@

\begin{table}[h!]
  \begin{center}
    \scalebox{0.7}{
    \begin{tabular}{|l|r|r|r|}
      \hline
Benchmarks        &base(-Ofast) &\GCM{}(-Ofast) &GCM/base\\\hline
400.perlbench	  & 2539	& 2493	& 0.98 \\\hline
401.bzip2	  & 718	        & 707	& 0.98 \\\hline
403.gcc	          & 5782	& 5722	& 0.99 \\\hline
429.mcf	          & 14	        & 17	& 1.21 \\\hline
433.milc	  & 635	        & 642	& 1.01 \\\hline
444.namd	  & 3223	& 3274	& 1.01 \\\hline
445.gobmk	  & 2166	& 2173	& 1.00 \\\hline
447.dealII	  & 11074	& 10234	& 0.92 \\\hline
450.soplex	  & 1122	& 1125	& 1.00 \\\hline
453.povray	  & 4701	& 4692	& 1.09 \\\hline
456.hmmer	  & 1197	& 1274	& 1.06 \\\hline
458.sjeng	  & 168	        & 176	& 1.05 \\\hline
462.libquantum	  & 118	        & 118	& 1.00 \\\hline
464.h264ref	  & 3379	& 3454	& 1.02 \\\hline
470.lbm	          & 33	        & 33	& 1.00 \\\hline
471.omnetpp	  & 524	        & 527	& 1.00 \\\hline
473.astar	  & 180	        & 201	& 1.12 \\\hline
482.sphinx3	  & 674	        & 670	& 0.99 \\\hline
483.xalancbmk	  & 5102	& 4965	& 0.97 \\\hline
Grand Total	  & 43349	& 42497	& 0.98 \\\hline
    \end{tabular}}
  \end{center}
  \caption{Number of spills before and after \gcm{}(\GCM{}) on SPEC2006 at
    -Ofast. Lower is better in \%age change. Spills reduced by 2\%.}
  \label{tab:spills}
\end{table}

Spills are listed in Table~\ref{tab:spills}. Spills were reduced by an average
of 2\% on the \SPEC{} benchmark suite. This is a result of cost-model which
limits the \gcm{} to those candidates which do not increase the register
pressure (except for loads Section~\ref{subsec:cost-models}) overall spills
still decreases.
